\documentclass[12pt]{memoir}
\title{Does Music Move?}
\author{}   
\date{}
\begin{document}
\maketitle

\subsection{Introduction}

Does music move?

Well, yes---if by ``music'', we meant the
vibrations of sound that allow us to hear what a
musician plays: from instrument to air to ear.

Or yes again---if by ``move'', we meant the way
we can indeed be \emph{moved} by music: from
table-taps to tangos to tears.

But the question is harder to answer, if we ask it
of music's movement in a more elementary sense.
---The sense in which we might say, of a rhythm,
that it quickens and slows; or of a melody,
that it rises and falls; or of a harmony, that
it departs and returns. And suppose we say all
this about the first movement of a symphony, not
thinking twice about calling this a ``movement''.
For we talk as if we hear just that in the
music---movement---and as if any piece of music
indeed moves itself in moving us.

Our talking this way has a point. For if we didn't
hear music move, would we hear it at all? Without
movement, music would seem no more than a series
of sounds.\footnote{For further discussion of
this, see Chapter VII, ``The Paradox of Tonal
Motion'', in Victor Zuckerkandl's \emph{Sound and
Symbol: Music and the External World}; and the
section ``Movement'' of Chapter 2, ``Tone'', in
Roger Scruton's \emph{The Aesthetics of Music}.}
But there is a problem with our talking this way,
despite its point. And this is the problem I
discuss in my lecture tonight. In its first part,
I explain what I take the problem to be. In its
second part, I explain why I take the problem
to be important. In its final part, I offer two
solutions to the problem, in the attempt to deepen
our sense of it.

\subsection{The nature of the problem.}

So what is the problem with our talking as
if music moved? The problem, in a word, is
\emph{space}: the space in which music moves, when
we hear it as music. ---The space that allows it
to quicken and slow, rise and fall, depart and
return. For this space makes something close to
complete sense on the one hand, yet something
closer to nonsense on the other.\footnote{For a
defense of the necessity of thinking that music
moves in space, but that the space in question
is metaphorical, see Chapter 4, ``Movement'', in
Roger Scruton's \emph{Understanding Music}.}

\subsubsection{The sense it makes.}

To see the sense it makes, we can start by
comparing the movement we hear in music, with
the variation we hear in sounds more generally.
Consider, for example, the difference between a
melody and a siren. Both involve a change in pitch
over time. And we might say, in that respect, that
both rise and fall. But unlike the siren, the
melody does this in a kind of articulated space,
usually conceived as a scale. And this space gives
us the impression of movement, from one place to
the next. We sense a change of place, in other
words, \emph{within} the melody's change in pitch,
as if having caused it.

As illustration, suppose we hear the beginning
of ``Twinkle Twinkle, Little Star'' on middle-C,
where C is followed by the G a fifth above.
Within this change in pitch, we hear a change of
place---from the first degree of a melodic scale
to the fifth; or otherwise put, from the $\hat{1}$
to the $\hat{5}$. And this change of place gives
us the impression of a movement having caused the
change in pitch: in this case, of a leap from
the $\hat{1}$ to the $\hat{5}$, causing C to be
followed by G.

Now suppose we hear the siren rise from C to G. We
don't hear this rise happen in the articulated
space of a scale; and indeed the continuity of the
rise would seem to preclude it, since the places
in a scale are discrete. But this means we are
given no impression of movement by the siren's
change in pitch, from one place to the next. It's
as if the siren rises only in time, not in space.
True, we can see, or at least infer, a change of
place \emph{behind} the siren's change in pitch,
in the fire engine, say, that produced it. But we
don't hear any change of place \emph{within} the
change in pitch, that might replace the fire
engine as cause, and turn what we hear as a signal
into what we could hear as a melody.

Of course, we may not know about a musical scale,
to account for the change of place we hear in
the melody. But this ignorance is only more
evidence for the sense made by the space of
music's movement, in its \emph{own} terms. And it
reflects a striking fact about what it means to
understand music. For we can develop an altogether
discerning musical ear, while remaining all but
illiterate about what we hear. And becoming
literate, by studying music theory, underlines
the sense music makes even without this theory.
For the topography of this theory---such as
the melodic scale---is more discovered than
invented, and in the discovery, more inhabited
than observed. One sign of this is the way we are
moved by music; for we are thus moved inwardly,
in the space we inhabit as selves. But I will
now try to show this more explicitly. And my
conclusion will be that it is because we 
\emph{inhabit} what we hear, in hearing music
move, that the space of this movement makes
complete sense. The completeness of this sense has
to do with the experience of inhabitance.

To begin to see this, recall the terms I used for
the melody ``Twinkle, Twinkle.'' I identified its
first two pitches by note---middle-C and the G a
fifth above. This identification depends on two
topographical facts in our perception of notes,
even outside a musical context. The first such
fact is that we hear a difference in pitch,
whereby one note is distinguished from another, as
a difference in relative place: one note sounding
higher, the other note lower. This perception is
also transitive: if one note sounds higher than a
second, and the second higher than a third, then
the first will also sound higher than the third.
Our perception of pitch difference so gives every
note its own position along an axis of height.
Hence my talk of the G a fifth \emph{above}
middle-C, to distinguish it, say, from the C a
fourth below.

Then there is my reference just now to the two G's
on either side of middle-C, along with my
reference to middle-C itself, in order to
distinguish that C from every other C there is.
This repetition of note letters reflects the
second topographical fact in our perception of
notes: the phenomenon of the octave. For while we
distinguish notes by their difference in pitch,
this difference reaches a kind of limit at the
interval called the ``octave'', where the notes
sound the same, despite their difference in pitch,
and are thus given the same letter as name.
Exactly \emph{why} we hear this sameness is hard
to say: Aristotle attributed it to a perception of
the whole number ratio two-to-one; while Victor
Zuckerkandl calls it a miracle.\footnote{Aristotle
    attributes the cause of the octave to the
    ratio 2:1 in Book II, Chapter 3, of his
    \emph{Physics}, 194b28--29. Zuckerkandl calls
    it a miracle in his discussion of the octave
    in the section ``Scale'' of Chapter VIII,
    ``The True Motion of Tones'', in \emph{Sound
        and Symbol}.} But however it happens, the
sameness we hear in notes, once their difference
in pitch reaches the octave, in effect contains
that difference. If we pass beyond it, we don't
encounter new notes, but only new instances,
higher or lower, of old notes. The octave thus
turns the axis of height, along which notes are
arranged by pitch, into a kind of circumference,
which continues to trace their increase or
decrease in pitch without end, but always to the
same place again.

Yet this image, of notes now arranged into a
circumference by the octave, is not yet an
image of the space in which music moves. For it
only comprehends the change in pitch involved,
as we might conceive this change to carry us
\emph{along} the circumference. But in hearing a
melody, we again hear 
a change of place \emph{within} the change in
pitch, such as the leap from the $\hat{1}$ to the
$\hat{5}$ at the start of ``Twinkle, Twinkle''.
And this description reflects another pair of
topographical facts in our perception of notes,
once we hear these notes in a musical context.
The first such fact is that we hear movement from
note to note, without having to hear any notes
between. So in ``Twinkle, Twinkle'', we hear a
leap from the $\hat{1}$ to the $\hat{5}$, without
having to hear the $\hat{2}$, $\hat{3}$, and
$\hat{4}$ first. And to hear them first would
not complete the leap, as if to fill it in, but
rather transform the leap into a climb. Why this
is a fact of musical perception may be as hard to
explain as the octave. But as a phenomenon, it
seems to involve a sense of being \emph{oriented}
in the movement we hear from note to note. It's 
as if we faced the note we were moving to, and
reached it as a goal, heedless of any notes on the
way.

The sense of being oriented among notes is even
clearer in the second topographical fact, which
involves our perception of notes in musical
contexts we call \emph{tonal}. It is this fact
that forced me to shift from letters to numbers,
when identifying the leap in ``Twinkle, Twinkle''
from the $\hat{1}$ to the $\hat{5}$. For in tonal
musical contexts---and the name ``tonal'' is
derived from this fact---we will hear one note as
a kind of center, which \emph{orients} us with
respect to the other notes we hear, as if to
provide a place from which to face them. The
central note is so assigned the number ``1'' in
analysis, and the other notes assigned other
numbers in reference to ``1''. Again, why we can
hear a certain note as central is hard to say. But
the phenomenon, as Zuckerkandl would remind us, is
dynamic, in an orientation more felt than seen. We
hear the central note as central, that is, by
sensing a stability in it relative to the other
notes, as if it provided a place to face them as a
center of gravity. This second topographical fact
so informs the first: for we then move from note
to note as if under the influence of a
gravitational pull, requiring effort to overcome,
and supplying momentum in success, in a deepened
sense of having faced the note being moved to, and
reaching that note as a goal.

If we accommodated these facts of musical
perception into the earlier image of a
circumference of notes, arranged by pitch and
bounded by the octave, then we might describe it
like this: we hear the movement in a melody, as in
music more generally, as if we were \emph{inside}
that circumference. For once we hear one note in
the melody as central, especially if we hear it as
a center of gravity, it's as if that note were
projected \emph{inside} this circumference from
its place along it. And this projection allows us
to move from note to note not simply \emph{along}
the circumference, through every note between, but
now \emph{across} the circumference, guided by the
one note inside it as a center of orientation.
This image so gives a geometric form to the
complete sense made by the space in which music
moves. For this is the sense in which we
\emph{inhabit} that space.

To be sure, the development of this sense, as
shown by the specifically \emph{tonal} context in
which we hear one note of a melody as central,
depends on the like development of a specific
form of musical art---the art of tonality, to
which we owe the music of the West. But there
would be no such art to develop, unless the result
made a difference to what we could hear; and in
this case, to what we could inhabit in what we
hear. And much of tonal music's development can be
explained as a deduction from the features of a
place we inhabit.

I spell this thought out briefly, in one example,
for those familiar with this development. If
we inhabited the space of music's movement, it
seems we should not be simply fixed at a single
center of orientation. We should rather be able
move that center, carrying it with us from place
to place. And move it we can---once tonal music
developed the device of modulation, to carry
us from key to key. We might also expect the
movement of that center to happen along an axis
of depth, away-and-back, not just to distinguish
it from the up-and-down movement between notes
along an axis of height, but also from a sense
of perspective, which implicitly belongs to our
sense of orientation when we inhabit a place.
And so we do move away and back---once tonal
music developed the harmony out of polyphony that
modulation relies upon. We might further expect
the movement of the center to clarify our sense,
as I described this above, of the gravity felt at
work in such a space. And clarified it is---once
the use of modulation effectively reduced the
modes of chant to the major-minor scale. We
might finally expect the movement of the center
to deepen our sense that we inhabit \emph{one}
space, which \emph{contains} the places moved
between, rather than many spaces distinguished and
divided by those places. And deepened it is---once
the use of modulation forced upon the tuning
of a scale the leveling of equal temperament;
for this restricts the notes of \emph{any} key to the
notes in \emph{every} key, and thereby places
``the whole of tonal space,'' as Roger Scruton has
put it, ``within reach of its every
occupant.''\footnote{Roger Scruton, \emph{The
Aesthetics of Music}, p.~244.}

Along with this deduction comes a plausible
measure of the greatness in a musical work.
The greater it is, the deeper it carries us into
the space of music's movement. And by that measure
of inhabitance, the best demonstration of the
complete sense this space can make, is found not
in the account I have just given of it, but rather
in those masterpieces of music---such as
Beethoven's \emph{Eroica}---where this space is
explored to a kind of limit, in the conquest of
it.

\subsubsection{The nonsense it makes.} But as I
said, the space of music's movement also makes a
kind of nonsense. To see this, we can start by now
comparing the movement we hear in music, with
motion as we observe this more generally.
Consider, for example, the difference between the
movement of a melody, like ``Twinkle, Twinkle''
again, and the motion of our hand in following
that melody, as if to conduct it. We could say
that both rise and fall, and do so not just in
time but in space, through a discernible change of
place. But there is a difference.  For as I
mentioned earlier, we hear the change of place in
the melody, such as the leap from the $\hat{1}$ to
the $\hat{5}$ in ``Twinkle, Twinkle'', without
having to hear any places between. The movement
happens, so we might say, \emph{discretely}. By
contrast, we see the hand's change of place happen
\emph{continuously}, from place to place through
\emph{all} the places between. And this continuity
would seem a necessary feature of its motion. For
if our hand got from place to place discretely,
like the melody did, skipping places along the
way, then it would look to us as if our hand
reached each place not by motion, but rather by
magic.  Or at least we'd be tempted to think there
was something in the space where the motion
occurred, beyond \emph{just space}, that was
interrupting the motion, disrupting the continuity
it would otherwise have. But this is one reason,
then, to think there isn't really a space for the
melody to move in. For if there were, then it
would allow the melody to move in it continuously
rather than discretely.

But this reflects another difference---involving
identity rather than continuity---between the
movement of the melody and the motion of our hand
in following it. For the melody is \emph{composed}
of the notes it moves between, while our hand is
not composed of the places it moves between. And
this explains at once why the melody has to move
discretely. For it has \emph{to become} what it
is. And this means passing through only those
notes that compose it, and that distinguish it,
thus composed, from any other melody. But then the
melody can only \emph{be} what it is by becoming
so, over an interval of time. And in this sense,
the melody is temporal rather than spatial, with
an identity in time rather than space. But this is
then another reason to think that there isn't
really a space for the melody to move in.  For if
there were, then it would allow the melody to
possess an identity in space; that is, it
would allow the melody to \emph{be} what it is in
space without having to become so, and to remain
what it is, unchanged, through every change of
place.

But here is perhaps a stronger way to put the
nonsense: If there really were a space for the
melody to move in, then there would be a melody
to hear, in the space where we hear it move.
But there isn't. We hear the leap in ``Twinkle,
Twinkle'', for example, without hearing anything
making the leap. For we hear this leap being
made between unleaping notes, and hear nothing
further to which we might attribute the leap. And
this seems true for music in general. We hear a
movement being made between unmoving notes, and
nothing further to which we might attribute the
movement. So we hear movement, but nothing making
the movement. Yet how, in that case, could there
\emph{be} any movement to hear? And what could be
making it?

Yet this is only the start of the nonsense. And
the end of it implicates our very inhabitance of
the space in question. We can see this through an
objection to the analysis I just gave. True, the
objection runs, we have a sense of the space in
which motion ordinarily occurs, that we cannot
apply to the movement of music. For applying it
makes nonsense of the movement, by depriving
this movement of any object to which it might be
attributed. But what follows from this? Perhaps
only that the space in which music moves, is
not ordinary, but extraordinary. And this is
one way to understand my earlier defense of the
sense---the complete sense---made by this space.
For this was not an observed sense of space, but
rather an inhabited sense of space. The sense, for
example, in which we \emph{face} the note being
moved to, and reach it as a goal. So if there
\emph{is} a space for music to move in, which
makes sense of the movement, then this space will
have to be conceived from within, as a matter of
inhabitance, rather than from without, as a matter
of observation.

Well, fair enough. But this makes the nonsense
of such a space even clearer. For what is it,
finally, that makes an inhabited sense of space
extraordinary rather than ordinary? Here is one
answer---the answer, for example, that Heidegger
might give.\footnote{For a further account of
Heidegger's actual answer, see Part One, Division
One, Chapters II and III of his \emph{Being
and Time}.} An inhabited sense of space is
extraordinary, in not stopping short of totality.
That is, our inhabited sense of space implicitly
includes everything there is, known or unknown;
anything captured in the word ``Being''. For it
is beings, finally, that we take ourselves to be
surrounded by, and our own being that provides
the place from which to face them. This is why,
despite our sometimes talking as if there were
more than one world, and even more than one world
we might inhabit, we can also talk intelligibly,
in a simpler yet deeper way, of \emph{the}
world, as if there were only one. And it is our
being-in-the-world, on this answer, that grounds
our inhabited sense of space.

But what happens, then, when our inhabited sense
of space is divided? ---For example, between
sleeping and dreaming, where it seems we inhabit
two spaces at once? We resolve the division,
evidently, by conferring worldhood on only one
such space, taking it \emph{to contain} the other
such space. That is, we credit only one 
space with the totality, and thus the reality, of
inhabitance, and regard the other space as merely
part of this totality. The credit we give to its
own totality, then, is the credit we give to a
dream. We still inhabit the dream---indeed more
alertly than we inhabit the bedroom in which we
dream, and often with a sense that everything is
put at stake in the dream, in a matter of life
or death. But in waking from the dream, we prove
it to be part of a larger space of inhabitance.
And this gives the dream's apparent totality the
status of mere appearance; and our inhabitance of
it, the form of an illusion. To our relief, or
perhaps our regret, what happened to us in the
dream, didn't really happen after all.

But if this is so, then it gives us reason to
suspect the very same thing of the space we
inhabit when hearing music move. The space may
well be illusory, making all the sense---but also
all the nonsense---of a dream. And in that case,
the greater the work of music, the deeper it
carries us into the dream. We hear the sound of
Beethoven's \emph{Eroica} surge forth in a concert
hall, seeming to make the whole world shake. Yet
the musicians barely move by comparison, while
the notes they play move not at all. And we
concert-goers stay glued to our seats---entranced.
And once the work is finished, in a triumph of
conclusiveness, we are released from the trance
in a daze---and the desire, perhaps, to have
remained. For we leave the concert hall likely
finding the world we truly inhabit unchanged by
what we heard, and nothing comparable to its
conclusiveness in the life we have to live.

This, then, is what I take to be the problem with
our talking as if music moved. Talking that way
fails to distinguish our experience of music from
a dream, in which nothing we experience really
happens.

\subsection{The importance of the problem.}

I also take this problem to be important. To
explain why, I discuss this problem again, but
now as a problem not simply with our experience
of music, but more generally, with our experience
of the world. For we talk not just as if music
moved, but as if \emph{anything} moved. Yet this
too involves a kind of nonsense, known since the
time of Parmenides. And it proves to be the same
kind of nonsense we encountered in the movement of
music.

To see this, recall my earlier analysis of why
there really isn't a space for a melody to move
in. For the melody, in this respect, is temporal
rather than spatial. It is composed of the places
it moves between, and can only be what it is by
becoming so. So there isn't a space, strictly
speaking, for the melody to move in, with the
continuity or identity of truly spatial things,
like our hand in following that melody.

Fair enough. But while our hand may be spatial
in this respect, the \emph{motion} of our hand
is temporal, just like the melody. In a sense,
it too is composed of the places moved between,
and can only be what it is by becoming so. But
this implies, on the earlier analysis, that there
isn't really a space for our hand to move in, or
indeed for anything to move in. There is only a
space for it to be contained in, and to occupy,
over the course of its motion, at every place
composing that motion. Space, in other words,
can only contain the path of the motion, not the
motion itself.

But here again is perhaps a stronger way to put
the nonsense. If there really were a space for
our hand to move in, then there would be a hand
to see, in the space where we see it move. But
there isn't. All we see is the motion being
made from place to unmoving place, by something
that occupies each such place. We see nothing
further, to which we might attribute \emph{the
motion} between places, rather than just \emph{the
occupancy} of those places. So we see motion, but
nothing making the motion. Yet how, in that case,
could there \emph{be} any motion to see? And what
could be making it?

One answer, the kind Zeno might give, is that
there \emph{isn't} any motion to see.\footnote{For
further discussion of Zeno's arguments against
motion, see Aristotle's \emph{Physics} Book VI,
chapters 2 and 9, and Book VIII, chapter 8; and
Lecture VI, ``The Problem of Infinity Considered
Historically,'' in Bertrand Russell's \emph{Our
Knowledge of the External World}.} When we think
we see something move, all we really see is
that it occupies different places at different
times. And the appearance of motion in this, is
something like the appearance of motion on a movie
screen, or in a flip book, from a succession of
images, each of a single place and time, that
happens too rapidly for us to detect. But this
first answer leads to a second answer, found for
example in calculus, where there \emph{is} a
motion to see.\footnote{For a further discussion
of this answer, see Lecture V, ``The Theory of
Continuity,'' and Lecture VII, ``The Positive
Theory of Infinity,'' in Russell's \emph{Our
Knowledge of the External World}.} When we see
something move on this answer, all we indeed see
is that something occupies different places at
different times. Still, it can be proved that
there are more such places---infinitely more, in
fact---than we could ever arrange from one to the
next. The places are many enough, then, to form a
\emph{real} continuum, beyond the mere appearance
of one in a rapid enough succession of places
arranged one to the next. And this makes for a
real continuity in the motion we see, giving it
all the reality it needs.

There is a sense, however, in which this second
answer misses the point. For the continuity in
question belongs to the motion's path rather
than to the motion itself. And the way this path
is proved to form a continuum---for example, in
Dedekind---is by analogy to a line, undivided
by time, where every place upon it is present
at once.\footnote{Dedekind draws the analogy in
Chapters II and III of his \emph{Continuity and
Irrational Numbers}.} But it is only when this
line is divided by time, where only one place upon
it is ever present at once---the point right now,
so to speak---that it seems we have the proper
analogy for the continuum of the motion, rather
than just its path. But how can a line divided
by time be proved continuous, if it is made of
only one point at a time?

The second answer so leads to a third answer, more
philosophical than mathematical, and found, for
example, in Bergson.\footnote{See, for example,
Bergson's lecture ``The Perception of Change'' in
the collection \emph{The Creative Mind}. There
Bergson even connects his account of motion
generally to the movement in a melody. Zuckerkandl
follows Bergson's lead in his own account of
music's movement in \emph{Sound and Symbol}.} This
answer is distinguished from the first answer in
taking the motion we see to be real, and from the
second answer in taking this reality to involve
more than just the continuity of the motion's
path. But the promise of this answer comes at a
price: for it embraces the nonsense that makes the
problem a problem. On this answer, that is, there
is indeed nothing to see, in the space where we
see anything move. And this is because the space
where we see anything move, is the space where the
motion belongs to us, rather than to anything we
might see outside us. So there \emph{is} a motion
to see, when \emph{we} are the ones making it.

On this answer, in other words, motion occurs in
the space we inhabit, rather than in the space
we observe. For in the space we observe, we see
only the continuous path of a motion, occupied
at every place by the thing that moves. But in
the space we inhabit, we see the motion itself,
which is not simply continuous, but indivisible.
The motion in this indivisible sense stretches
from the beginning of its path to the end in a
single bound, as if the entire path were a single
place for the moving thing to occupy. And we know
this indivisibility when moving ourselves. For
in that case, we face the end of our motion at
the beginning of it, and reach the end as a goal,
heedless of any places on the way---just like in
the movement of a melody. And the path we trace
in reaching that goal can be divided only by our
changing the goal, in a motion now different from
what it was, but again indivisible. This too is
like the movement in a melody; for only further
notes can divide the distance from note to note,
but thereby produce a new melody out of the old
one, showing the movement between notes in any one
melody to be, indeed, indivisible.

But what should we make of this answer? It makes
the same appeal to our inhabited sense of space
that we encountered before in making sense of
the movement in music, only now to make sense of
motion as such. But does it have the same problem?
That is, can we show, or know, that we aren't
dreaming when we see anything move?

The question put this generally might seem absurd.
After all, to suppose we \emph{were} dreaming
when seeing anything move, would mean suspecting
something illusory about our inhabitance of
\emph{the world}, and not just of so-called other
worlds, such as dreams in the ordinary sense, or
music, that we might plausibly find illusory in
their proving \emph{contained} in the world, mere
parts of that totality.

But there is reason still, to suspect something
close to illusory about our inhabitance
of the world. And it explains why a question
like ``Are we dreaming right now?'' has
been raised in philosophy---for example, by
Descartes.\footnote{Descartes raises this question
in his \emph{First Meditation}. It is also
discussed in Plato's \emph{Theaetetus}, 158b-d;
and mentioned, but also dismissed, in Aristotle's
\emph{Metaphysics}, 1011a6.} For while we may
not be dreaming when we see anything move, there
is arguably nothing in our inhabitance of the
world to prove it. And we lack this proof, so the
argument goes, precisely because we inhabit the
world. If we had the proof, we would no longer
inhabit the world, but merely observe it.

To see why this argument is worth considering,
consider, first, a striking fact about optical
illusions, such as the look of a stick
half-plunged in water, and seemingly half-bent
by it. Our knowing the stick stays straight does
nothing to dispel the illusion. But why? One
answer is that there is nothing \emph{in} the
illusion to \emph{tell} us it's an illusion, and
if there were, then the illusion wouldn't be an
illusion. We so remain \emph{in} the illusion, or
inhabit it, as a matter of perception, even if we
stand outside the illusion, or merely observe it,
as a matter of knowledge.

Something like this is also at work in dreams,
taking ``dreams'' again in the ordinary sense.
For our dreams so often contain implausible or
even impossible events, not to mention the sense
in which all its events, once we wake up, prove
unreal. So why did they seem so real in the
dream? Again, one answer is that there is nothing
\emph{in} the dream, no matter how implausible or
impossible, to tell us we are dreaming, and if
there were, then we wouldn't be dreaming. So we
inhabit the dream rather than merely observe it.
It is only when something in the dream wakes us up
from it, and puts us outside the dream to observe
it, that we can know it for the dream that it is.

We can give this same answer to the question of
why our experience of music would still seem so
real, even if we were convinced---say, by my
lecture tonight---that it was only a dream. For
there is nothing \emph{in} the music to tell us
so, even if there is some account of the music to
tell us so. And again, this is because we hear
music as music only by inhabiting it, past the
point of any account we might give of it.

This answer, then, points to perhaps the most
decisive, if negative, feature of our inhabitance
of the world. There is nothing we might encounter
in the world---motion, for example---to tell us
the encounter is illusory. And if there were, then
we wouldn't be having the encounter. So for all
we know, we might be dreaming. Arguing against
this prospect, to be sure, is our sense that this
is indeed the world that we inhabit; our sense,
in other words, that we are surrounded by beings
rather than seemings, and that we encounter these
beings from a being rather than seeming of our
own. But this sense can never be decisive. For
the cost of inhabiting the world, on this answer,
is to be past the point of knowing whether this
inhabitance is only a dream.

This, then, is what I take to be the importance of
the problem with our talking as if music moved.
For it illustrates a more general problem, with
our talking as if anything moved. Talking that
way fails to distinguish our experience of the
world from a dream, in which nothing we experience
really happens.

\subsection{Conclusion}

I conclude my lecture with two solutions to this
problem, in the attempt to deepen our sense of it.
The first solution might be called Cartesian, or
scientific; and it is found implicitly in how we
talk when trying to \emph{explain} the world we
experience. For then we talk as if our inhabitance
of the world is indeed a dream, from which there
is no waking up. In other words, the terms in
which we explain the world, as a matter of
observation, are not the terms in which we
encounter the world, as a matter of
inhabitance---and in many cases, such as quantum
mechanics, could *never* be the terms of such an
encounter. Yet we take this discrepancy as
a sign that the world has indeed been explained rather
than erased. This is why we look for an
explanation of our encounter with music, not in
Beethoven's \emph{Eroica}, but rather in the
physicist's account of wave phenomena, or the
neuroscientist's account of brain patterns, or the
biologist's account of evolutionary adaptations,
or even the Nietzschean's account of a sickness in
the soul. Despite the fact that we don't encounter
music as music in such terms, we credit these
accounts as attempts to explain that encounter.
And why? Because we find nothing self-justifying
in the encounter itself: it might well be only a
dream. And to solve this problem, we explain the
dream away. That is, we explain the encounter by
avoiding any appeal to the sense the encounter
makes only from a place of inhabitance---and often
by challenging that sense.

This brings me to the second solution, which might
be called Heideggerian, or metaphysical. In this
solution, not having a proof that our inhabitance
of the world is real, is all the proof for it that
we need. There is nothing to tell us we might be
dreaming, true; but this is because we aren't.

This solution already sounds unpromising. After
all, there is nothing in \emph{a dream} to
tell us we are dreaming, but we are. On this
solution, however, it turns out we aren't, at
least as a matter of inhabitance. For any space of
inhabitance is real, on this account, precisely
because it can be inhabited. And this means we are
\emph{always} surrounded by beings rather than
seemings, and always encounter them from a being
rather than seeming of our own, whether we find
ourselves in a dream, in a symphony, or in the
world that contains them. True, we have reason to
think otherwise, again since these smaller spaces
are contained in a larger space that proves their
own totality to be illusory. But this assumes
that any space of inhabitance lies, indeed, in
space. Yet on this solution, it is only a space
of observation that lies in space. Any space of
inhabitance lies instead in time.

Perhaps this solution now sounds even more
unpromising. For how can there be any room in time
for a space of inhabitance, when the only part of
time that ever exists is `right now', that instant
between the no-longer and not-yet at the seeming
size of a point? Still, this very fact about
time's evanescence turns the illusion of motion
into an object of wonder if not astonishment.
For according to our earlier analysis, motion is
what it is, only in time. But if time exists only
at a point, then how could we ever encounter a
temporal whole like motion---even in the form of
an illusion? How could we even dream that we hear
the rise and fall of a melody, or see the rise and
fall of our hand in following it?

It is in offering an answer to this question that
the solution starts to look more promising. And
the answer is this. `Right now' may be no larger
than a point. But this is still room enough to
establish a space of inhabitance, in providing a
center of orientation. And this is the center from
which the not-yet lies ahead of us, the no-longer
lies behind us, and the right-now is always
with us. And in being always at this center,
we are always in the world, even when we find
ourselves in a dream, or a symphony. We inhabit
even these smaller spaces from the same center of
orientation, where the not-yet is ahead of us as
a matter of expectation, the no-longer is behind
us as a matter of memory, and the right-now is
always with us as a matter of attention. It is in
these terms, temporal terms, that any space of
observation becomes a place of inhabitance.

It is also in these temporal terms that any place
of inhabitance is real rather than illusory. For
this center of orientation, despite being a mere
point in time, comprehends a totality of time
for each of us: from a beginning in birth that
no one else can share, to an end in death that
no one else can know, in a span of time without
remainder. This is the center, then, from which
the not-yet ahead of us is a future to face; the
no-longer behind us is a past to bear; and the
right-now always with us is a present at stake.
So in being always at this center, at a point of
totality as it were, we are always in a world that
is real, even when we find ourselves in a dream
or symphony. We inhabit even these smaller spaces
from the same center of orientation, with a future
to face, a past to bear, and a present at stake.
This is why we can be released from their spell,
without having recovered from it. And in this
sense, again temporal rather than spatial, we are
not carried from the world by a dream or symphony,
but concentrated in it.

Or to summarize this solution to the problem in
conclusion: we can only talk as if \emph{anything}
moved, much less as if music moved, insofar
as the space we inhabit in talking that way,
makes complete sense of what we say, without
any room for nonsense. And this is the space we
inhabit right now: the space made possible by
time.\footnote{This is a lecture I gave at the
Annapolis campus of St.~John's College on February
15, 2013. I would like to thank Gabriela Hopkins
for helping me improve the final draft with
comments and conversation about earlier drafts.
---Daniel Harrell}
  
  \end{document}
